---
title: "Programming Neural Networks"
author: "Dustin Pierce"
date: "2024-05-17"
output: pdf_document
---

Below are some options for programming your own Neural Networks. You ARE ALLOWED to use AI or external resources to assist you. Please give credit (cite).

## Keras

[keras](https://tensorflow.rstudio.com/guides/keras/basics) is a Neural Network library. `Keras` is fast, flexible and provides lots of functionality. The installation process is labor-intensive, but is worth the effort. 

SEE [here](https://docs.google.com/document/d/1-q86kpyCACWcYWHcmbd4XOkvRnif_kOZ/edit?usp=sharing&ouid=107235758751914785357&rtpof=true&sd=true) for the link to the installation instructions

```{r}
# Load the libraries (this assumes all packages installed)
library(tensorflow)
library(keras3)
library(tidyverse)

# Create a simple neural network model for testing
model <- keras_model_sequential() %>%
  # add hidden layer with 128 neurons and ReLu activation
  layer_dense(units = 128, activation = 'relu', input_shape = c(784)) %>%
  # add drop out (shut off random weights each iteration)
  layer_dropout(rate = 0.2) %>%
  # add output layer with 10 neurons with softmax activation
  layer_dense(units = 10, activation = 'softmax')

# Compile the model (define architecture)
model %>% compile(
  # optimize with Adaptive Moment Estimation
  optimizer = 'adam',
  # cost function is cross entropy (used for classification)
  loss = 'sparse_categorical_crossentropy',
  # measure accuracy of classification
  metrics = c('accuracy')
)

# Load the MNIST dataset (hand written digits)
mnist <- dataset_mnist()
# TRAIN => model learns from training data
x_train <- mnist$train$x
y_train <- mnist$train$y
# TEST => unseen data to vet model
x_test <- mnist$test$x
y_test <- mnist$test$y

# Preprocess the data (center the pixel data)
x_train <- array_reshape(x_train, c(nrow(x_train), 784)) / 255
x_test <- array_reshape(x_test, c(nrow(x_test), 784)) / 255

# Train the model 
history <- model %>% 
  fit(x_train, y_train,
      epochs = 5,
      batch_size = 32,
      validation_split = 0.2
)

# Evaluate the model
score <- model %>% evaluate(x_test, y_test)
print(score)


```


## Neuralnet 

[neuralnet](https://cran.r-project.org/web/packages/neuralnet/neuralnet.pdf) is an R package for training neural networks. Compared to `keras`, the `neuralnet` package is far less flexible, but the installation is instantaneous. 

```{r}
# Step 1: Install and load required packages
# install.packages("neuralnet")
library(neuralnet)
library(datasets)

# Step 2: Prepare the data
# Load the iris dataset
data(iris)

# We need to transform the species into numerical values
iris$Species <- as.numeric(factor(iris$Species)) - 1

# Split the data into training and testing sets
set.seed(123)  # For reproducibility
sample_index <- sample(seq_len(nrow(iris)), size = 0.7 * nrow(iris))
train_data <- iris[sample_index, ]
test_data <- iris[-sample_index, ]

# Normalize the data
maxs <- apply(train_data[, -5], 2, max)
mins <- apply(train_data[, -5], 2, min)
train_data[, -5] <- scale(train_data[, -5], center = mins, scale = maxs - mins)
test_data[, -5] <- scale(test_data[, -5], center = mins, scale = maxs - mins)

# Step 3: Train the neural network
# Define the formula for the neural network
formula <- Species ~ Sepal.Length + Sepal.Width + Petal.Length + Petal.Width

# Train the neural network
set.seed(123)  # For reproducibility
nn <- neuralnet(formula, data = train_data, hidden = c(5, 3), linear.output = FALSE)

# Step 4: Evaluate the model
# Predict on the test set
predictions <- compute(nn, test_data[, -5])$net.result

# Convert probabilities to class labels
predicted_labels <- apply(predictions, 1, function(x) ifelse(x > 0.5, 1, 0))

# Calculate accuracy
accuracy <- sum(predicted_labels == test_data$Species) / nrow(test_data)
print(paste("Accuracy:", accuracy))

# Plot the neural network
plot(nn)

```

## Custom Code

Using an external software package is not required, of course. We can write our own code or use AI to write custom code for a Neural Network. You may use `ChatGPT` or equivalent AI model to help you code your Neural Networks.
